Multivariate Ore Polynomials in SageMath

### 2019-05-17 papers/issac19demo/main.tex
inputenc
software demo: minor presentation fix
0.4

\documentclass[11pt]{article}
\usepackage[fleqn]{amsmath}
\usepackage{sigsam, sagetex, tikz}
\usepackage{hyperref}
\usepackage{booktabs}

@@

All operations discussed so far are generalizations of the univariate counterparts that already existed
in earlier versions. A feature that is inherently multivariate is creative telescoping~\cite{zeilberger91,chyzak00,chyzak14}.
Here, the input is a left ideal $I\subseteq A$, say $A=\set Q(x,y)\<Dx,Dy>$ and the task is to find
the ideal consisting of all operators $P\in\set Q(x)\<Dx>$ such that there exists $Q\in A$ with $P-D_yQ\in I$.
The operators $P$ are called \emph{telescopers,} and the corresponding operators $Q$ are called their \emph{certificates.}
Creative telescoping is a central operation for evaluating definite sums and integrals~\cite{chyzak14}.
We cannot explain here in detail the relevance of creative telescoping for these applications, but the use of the
package is easily demonstrated:
\begin{sageexample}
  sage: A.ideal([Dx - 2*x*y^2, Dy - 2*x^2*y]).ct(Dy)
\end{sageexample}
The output is a pair $([P_1,P_2,\dots],[Q_1,Q_2,\dots])$ whose first component is a basis of the ideal of telescopers
and whose second component contains the certificates corresponding to the basis elements.
The example above originates from the integration problem $\int_y\exp(x^2y^2)dy$. The following example is the essence
of a proof of the binomial theorem $\sum_k\binom nk=2^n$:
\begin{sageexample}
  sage: B.<Sx,Sy> = OreAlgebra(R)
  sage: B.ideal([(y+1)*Sy + (y-x), (x-y+1)*Sx - (x+1)]).ct(Sy - 1)
\end{sageexample}

@@

The examples shown above need virtually no computation time. In order to get some idea about the performance of our
implementation, we consider a set of 19 creative telescoping problems that appeared in a study of restricted lattice
walks~\cite{bostan16b}.
Each of the 19 problems starts from a certain rational function $r\in\set Q(u,v,t)$ and the task consists of two
applications of creative telescoping. Writing $I$ for the ideal of annihilating operators of $r$ in $\set Q(u,v,t)\<Du,Dv,Dt>$,
we first have to compute the ideal $T_v$ in $\set Q(u,t)\<Du,Dt>$ of all telescopers of $I$ with respect to~$Dv$ and then
the ideal $T_{u,v}$ in $\set Q(t)\<Dt>$ of telescopers of $T_v$ with respect to~$Du$. The smallest rational function in
the collection is
\[
\frac{{-}u^2 v^2{+}u^2{+}v^2{-}1}{t u^3 v^2{-}t u^3 v{+}t u^2 v^3{-}2 t u^2 v^2{+}2 t u^2 v{-}t u^2{-}t u v^3{+}2 t u v^2{-}2 t
  u v+t u-t v^2+t v-u^2 v^2+u^2 v+u v^2-u v},
\]
and the largest is about twice as long. The complete data is available in ore\_algebra.examples.ssw.

>>>>

\documentclass[11pt]{article}
\usepackage[fleqn]{amsmath}
\usepackage{sigsam, sagetex, tikz}
\usepackage{hyperref}
\usepackage{booktabs}
\usepackage[utf8]{inputenc}

@@

All operations discussed so far are generalizations of the univariate counterparts that already existed
in earlier versions. A feature that is inherently multivariate is creative telescoping~\cite{zeilberger91,chyzak00,chyzak14}.
Here, the input is a left ideal $I\subseteq A$, say $A=\set Q(x,y)\<Dx,Dy>$ and the task is to find
the ideal consisting of all operators $P\in\set Q(x)\<Dx>$ such that there exists $Q\in A$ with $P-D_yQ\in I$.
The operators $P$ are called \emph{telescopers,} and the corresponding operators $Q$ are called their \emph{certificates.}
Our new version of ore\_algebra includes an implementation of Chyzak's algorithm~\cite{chyzak00} for creative telescoping in the general D-finite setting.

Creative telescoping is a central operation for evaluating definite sums and integrals~\cite{chyzak14}.
We cannot explain here in detail the relevance of creative telescoping for these applications, but the use of the
package is easily demonstrated:
\begin{sageexample}
  sage: A.ideal([Dx - 2*x*y^2, Dy - 2*x^2*y]).ct(Dy)
\end{sageexample}
The output is a pair $([P_1,P_2,\dots],[Q_1,Q_2,\dots])$ whose first component is a basis of the ideal of telescopers
and whose second component contains the certificates corresponding to the basis elements.
The example above originates from the integration problem $\int_y\exp(x^2y^2)dy$. The following example is the essence
of a proof of the binomial theorem $\sum_k\binom nk=2^n$:
\begin{sageexample}
  sage: B.<Sx,Sy> = OreAlgebra(R)
  sage: B.ideal([(y+1)*Sy + (y-x), (x-y+1)*Sx - (x+1)]).ct(Sy - 1)
\end{sageexample}

@@

The examples shown above need virtually no computation time. In order to get some idea about the performance of our
implementation, we consider a set of 19 creative telescoping problems that appeared in a study of restricted lattice
walks~\cite{bostan16b}.
Each of the 19 problems starts from a certain rational function $r\in\set Q(u,v,t)$ and the task consists of two
applications of creative telescoping. Writing $I$ for the ideal of annihilating operators of $r$ in $\set Q(u,v,t)\<Du,Dv,Dt>$,
we first have to compute the ideal $T_v$ in $\set Q(u,t)\<Du,Dt>$ of all telescopers of $I$ with respect to~$Dv$ and then
the ideal $T_{u,v}$ in $\set Q(t)\<Dt>$ of telescopers of $T_v$ with respect to~$Du$. The smallest rational function in
the collection is
\[
\frac{{-}u^2 v^2{+}u^2{+}v^2{-}1}{t u^3 v^2{-}t u^3 v{+}t u^2 v^3{-}2 t u^2 v^2{+}2 t u^2 v{-}t u^2{-}t u v^3{+}2 t u v^2{-}2 t
  u v{+}t u{-}t v^2{+}t v{-}u^2 v^2{+}u^2 v{+}u v^2{-}u v},
\]
and the largest is about twice as long. The complete data is available in ore\_algebra.examples.ssw.
